---
title: "Getting started with bayesSYNC: a simulated biomarker example"
author: "bayesSYNC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with bayesSYNC: a simulated biomarker example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
has_splinets <- requireNamespace("Splinets", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  dev = "png",
  dpi = 150,          # higher resolution than the html_vignette default
  fig.retina = 2,     # crisp on high-density screens
  out.width = "100%",
  eval = has_splinets # skip the worked example if Splinets is absent
)
```

```{r load}
library(bayesSYNC)
```

## What this vignette shows

`bayesSYNC` implements a Bayesian functional factor model for high-dimensional
longitudinal data. It represents many observed curves (for example biomarker
trajectories) through a small number of shared, time-varying latent factors
that can be read as coordinated biological programmes or pathways. Each factor
loads on a subset of biomarkers through a sparse loading matrix, and its
temporal shape is described by functional principal component analysis (FPCA),
which splits within-factor variation into orthogonal modes ordered by the
variance they explain. Subject heterogeneity enters through low-dimensional
FPCA scores. The method is described in Jaoua, Temko and Ruffieux (2026),
*A scalable Bayesian functional factor model for high-dimensional longitudinal
molecular data*, [arXiv:2603.20844](https://arxiv.org/abs/2603.20844); section
numbers below refer to that paper.

This tutorial runs the whole workflow on a small simulated dataset, so every
estimate can be checked against a known truth. We use `generate_bayesSYNC_data()`
to produce the curves, then fit the model with a deliberately over-specified
number of factors and components and let the sparsity-inducing priors reveal how
many are actually supported by the data. This mirrors the recommended practice
of inferring the effective number of factors by over-specification (Section 4.3).

We follow `N` patients over time `t`, measuring `p` biomarkers at a handful of
irregular visits per patient. The model works on normalised time, so `t` must
lie in `[0, 1]`; Section 5 shows how to rescale an arbitrary time range and map
the results back. Here `p = 10` is a deliberately small toy so the vignette runs
quickly, but the method scales far beyond this: it has been applied to data with
up to 20,000 variables, matching the order of the number of human genes.
Building this vignette fits the over-specified model and takes a few minutes.

```{r splinets-note, eval = !has_splinets, echo = FALSE, results = "asis"}
cat("> **Note.** This vignette uses `generate_bayesSYNC_data()`, which depends on",
    "the **Splinets** package. Splinets is not installed, so the worked example",
    "below is not evaluated. Install it with `install.packages(\"Splinets\")` to",
    "run the tutorial.\n")
```

## 1. Simulate the data

`generate_bayesSYNC_data()` takes a mean function with signature
`mu_func(t, j, p)`, the numbers of subjects, biomarkers, factors and components,
and a per-subject count of visits. The Beta parameters `a_om` and `b_om` control
how many biomarkers load on each factor: the default `b_om = p` is very sparse,
so here we use a denser setting that gives each factor a clear block of loading
biomarkers, which keeps the illustration easy to read.

```{r simulate}
set.seed(1)

N <- 50                 # patients
p <- 10                  # biomarkers
Q_true <- 2              # true latent factors
L_true <- 2              # true FPCA components per factor

# mean function: note the required (t, j, p) signature
mu_func <- function(t, j, p) 1.5 * j + sin(2 * pi * t + j)

n_visits <- sample(3:10, N, replace = TRUE)   # irregular number of visits per patient

sim <- generate_bayesSYNC_data(
  N = N, n = n_visits, p = p, Q = Q_true, L = L_true,
  mu_func = mu_func,
  a_om = 5, b_om = 2,          # dense loadings: many biomarkers load on each factor
  vec_sd_zeta = c(1, 0.5),     # mode 1 carries more variance than mode 2
  vec_sd_eps = 0.2,            # measurement-noise standard deviation
  seed = 1
)

time_obs <- sim$time_obs
Y        <- sim$Y

# how many biomarkers actually load on each true factor
colSums(sim$gamma)
```

The outputs `Y` and `time_obs` are already in the format `bayesSYNC()` expects:
`Y[[i]][[j]]` holds patient `i`'s biomarker `j`, measured at the times in
`time_obs[[i]]`.

## 2. Fit an over-specified model

We deliberately ask for more factors and modes than the truth (`Q = 5`, `L = 5`)
and let the model prune what it does not need, following the over-specification
strategy for learning the number of factors (Section 4.3). `bool_scale = FALSE`
because the biomarkers already share a comparable residual scale;
`set_hyper(d_0 = 1)` gives a neutral inclusion prior. We keep the default
`maxit = 1000`. This is the heaviest step of the vignette because the cost grows
with `Q` and `L`; lower them for quicker iteration.

```{r fit}
Q_fit <- 5
L_fit <- 5

fit <- bayesSYNC(time_obs = time_obs, Y = Y, L = L_fit, Q = Q_fit,
                 bool_scale = FALSE, list_hyper = set_hyper(d_0 = 1),
                 n_cpus = 1,
                 tol_abs = 1e-3, tol_rel = 1e-4,   # tol_rel loosened so it converges within maxit
                 seed = 1, verbose = FALSE)

cat("iterations run:", fit$i_iter, "\n")
```

## 3. How many factors and modes are supported?

Two quantities drive the selection. The factor-level posterior probability of
inclusion `factor_ppi` gives, for each of the five fitted factors, the
probability that at least one biomarker contributes to it. We keep the factors
whose activity is high and treat the rest as empty. Within each retained factor,
the cumulated percentage of variance explained across its FPCA modes tells us how
many modes are needed; the remaining modes carry negligible variance.

```{r selection}
# (a) retain factors by their activity
round(fit$factor_ppi, 3)
active <- which(fit$factor_ppi > 0.5)
cat("retained factors:", length(active), "( expected", Q_true, ")\n")

# (b) within each retained factor, number of modes reaching 99% of the variance
n_modes <- sapply(fit$list_cumulated_pve[active],
                  function(cp) which(cp >= 99)[1])
n_modes                                   # expected 2 for each retained factor
```

```{r fig-selection, fig.width = 8, fig.height = 3.4}
op <- par(mfrow = c(1, 2), mar = c(4, 4.5, 3, 1))

barplot(fit$factor_ppi, names.arg = paste0("F", 1:Q_fit),
        ylim = c(0, 1), col = "skyblue2", border = NA,
        xlab = "Fitted factor", ylab = "Factor activity (PPI)",
        main = "Factor selection")
abline(h = 0.5, lty = 2, col = "grey40")

# cumulated variance explained within the retained factors
cp_mat <- sapply(fit$list_cumulated_pve[active], function(cp) cp[1:L_fit])
matplot(1:L_fit, cp_mat, type = "b", pch = 19, lty = 1,
        col = c("darkolivegreen3", "plum3"), ylim = c(0, 100),
        xlab = "FPCA mode", ylab = "Cumulated variance explained (%)",
        main = "Within-factor variance")
abline(h = 99, lty = 2, col = "grey40")
legend("bottomright", paste("retained factor", seq_along(active)),
       col = c("darkolivegreen3", "plum3"), lwd = 2, bty = "n", cex = 0.9)
par(op)
```

The left panel shows two factors clearly above the activity threshold and three
near zero, recovering `Q_true = 2`. The right panel shows the cumulated variance
climbing to essentially 100% by the second mode, recovering `L_true = 2`.

## 4. Align the estimates to the truth

Latent factors are identified only up to order, sign and an overall
loading/score scale, so before comparing estimates with the simulation we align
them. `match_factor_and_sign()` uses the true loadings and scores to permute the
fitted factors, keep the `Q_true` that match, flip signs and rescale. On real
data there is no truth to align to: you would stop at Section 3 and interpret the
retained factors directly.

```{r align}
aligned <- match_factor_and_sign(
  B = sim$B, B_hat = fit$B_hat, ppi = fit$ppi, factor_ppi = fit$factor_ppi,
  Zeta = sim$Zeta, list_Zeta_hat = fit$list_Zeta_hat,
  list_list_Phi_hat = fit$list_list_Phi_hat,
  list_cumulated_pve = fit$list_cumulated_pve,
  list_Cov_zeta_hat = fit$list_Cov_zeta_hat
)

B_hat_al    <- aligned$perm_B_hat              # p x Q_true
Phi_hat_al  <- aligned$perm_list_list_Phi_hat  # list of Q_true, grid x L_fit
Zeta_hat_al <- aligned$perm_list_Zeta_hat      # list of Q_true, N x L_fit
pve_al      <- aligned$perm_list_cumulated_pve

# true and estimated loadings side by side. We rename the columns so the two
# blocks are unambiguous: the fitted factors keep their original labels (e.g.
# Factor_4, Factor_5) after selection, which do not match the true factor order.
B_true <- sim$B
rownames(B_true) <- rownames(B_hat_al) <- as.character(1:p)
colnames(B_hat_al) <- paste0("est_F", 1:Q_true)
colnames(B_true)   <- paste0("true_F", 1:Q_true)
round(cbind(B_hat_al, B_true), 2)
```

## Results: recovering the shared temporal structure

The five panels below compare the fit (blue) with the simulated truth (black).
Together they show that the model recovers, from noisy and irregularly sampled
curves, the four ingredients of the generative model: the biomarker trajectories
themselves, which biomarkers drive each factor, the temporal shape of each
factor, and where each subject sits along those shapes.

```{r plot-setup}
time_g  <- fit$time_g
col_sim <- "black"
col_est <- "skyblue2"

# true noise-free curve for patient i, biomarker j, on the dense grid
true_curve_g <- function(i, j) {
  val <- sim$mu_g[[j]]
  for (q in 1:Q_true) {
    val <- val + sim$B[j, q] * as.vector(sim$Phi_g[[q]] %*% sim$Zeta[[q]][i, ])
  }
  val
}
```

### Reconstructed biomarker trajectories

Each panel is one patient and one biomarker. The blue curve is the fitted
trajectory with its 95% pointwise credible band; the grey dots are the actual
(noisy) observations; the black curve is the noise-free truth. The band should
cover the truth and the fit should track it even where visits are sparse, because
the model borrows strength across subjects and biomarkers.

```{r fig-trajectories, fig.width = 10, fig.height = 2.5}
set.seed(2)
pairs_i <- sample(1:N, 5)
pairs_j <- sample(1:p, 5)
op <- par(mfrow = c(1, 5), mar = c(4, 4, 3, 1), cex.main = 1)
for (k in 1:5) {
  i <- pairs_i[k]; j <- pairs_j[k]
  yh <- fit$list_Y_hat[[i]][[j]]; yl <- fit$list_Y_low[[i]][[j]]; yu <- fit$list_Y_upp[[i]][[j]]
  yt <- true_curve_g(i, j)
  plot(time_g, yh, type = "n", ylim = range(c(yl, yu, Y[[i]][[j]], yt)),
       xlab = "Time", ylab = "Biomarker value",
       main = sprintf("Subject %d, biomarker %d", i, j))
  polygon(c(time_g, rev(time_g)), c(yl, rev(yu)), col = adjustcolor(col_est, 0.2), border = NA)
  lines(time_g, yt, lwd = 2, col = col_sim)
  lines(time_g, yh, lwd = 2, col = col_est)
  points(time_obs[[i]], Y[[i]][[j]], pch = 19, col = "grey45", cex = 0.8)
  if (k == 1) legend("topleft", c("simulated", "estimated", "observed"),
                     col = c(col_sim, col_est, "grey45"), lwd = c(2, 2, NA),
                     pch = c(NA, NA, 19), bty = "n", cex = 0.85)
}
par(op)
```

### Loading matrix: which biomarkers drive each factor

The loading matrix is the map from biomarkers (rows) to factors (columns). A
strongly coloured cell means the biomarker contributes to that factor, red for a
positive loading and blue for a negative one; near-white rows are biomarkers the
factor ignores. This sparse structure is what makes each factor interpretable as
a pathway acting on a specific subset of biomarkers. The left panel is the truth;
the right panel is the estimate, annotated with the sign of the true loading so
position and sign can be checked at a glance.

```{r fig-loadings, fig.width = 7, fig.height = 5.5}
plot_loading_matrix <- function(M, main = "", sign_annot = NULL) {
  M[!is.finite(M)] <- 0
  pn <- nrow(M); qn <- ncol(M)
  m <- max(abs(M)); if (!is.finite(m) || m == 0) m <- 1
  pal <- colorRampPalette(c("#4a90c2", "white", "#e05a4d"))(101)
  v2c <- function(v) { if (!is.finite(v)) v <- 0; pal[round((v + m) / (2 * m) * 100) + 1] }
  plot(NA, xlim = c(0.5, qn + 0.5), ylim = c(0.5, pn + 0.5),
       xlab = "Factor", ylab = "Biomarker", axes = FALSE, main = main)
  for (j in 1:pn) { yv <- pn - j + 1; for (q in 1:qn) {
    rect(q - 0.5, yv - 0.5, q + 0.5, yv + 0.5, col = v2c(M[j, q]), border = "grey80")
    if (!is.null(sign_annot) && sign_annot[j, q] != "") text(q, yv, sign_annot[j, q], cex = 0.9)
  }}
  axis(1, at = 1:qn, labels = 1:qn, tick = FALSE, line = -0.5)
  axis(2, at = pn:1, labels = 1:pn, las = 1, tick = FALSE, line = -0.5, cex.axis = 0.8)
}
true_sign <- matrix("", p, Q_true); true_sign[B_true > 0] <- "+"; true_sign[B_true < 0] <- "-"
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot_loading_matrix(B_true,   main = "Simulated loadings")
plot_loading_matrix(B_hat_al, main = "Estimated loadings", sign_annot = true_sign)
par(op)
```

### Eigenfunctions: the temporal shape of each factor

Within a factor, FPCA decomposes the subject curves into orthogonal modes
ordered by the variance they explain. Each panel is one mode of one factor: the
black curve is the true eigenfunction, the blue curve the estimate, and the title
gives the share of within-factor variance carried by that mode. Mode 1 captures
the dominant temporal pattern, mode 2 a smaller higher-order fluctuation.

```{r fig-eigen, fig.width = 8, fig.height = 6.5}
per_comp_pve <- function(cp) c(cp[1], diff(cp))
op <- par(mfrow = c(Q_true, L_true), mar = c(4, 4.2, 3, 1), cex.main = 0.9)
for (q in 1:Q_true) {
  pv <- per_comp_pve(pve_al[[q]])
  for (l in 1:L_true) {
    yl <- range(c(sim$Phi_g[[q]][, l], Phi_hat_al[[q]][, l]))
    plot(time_g, sim$Phi_g[[q]][, l], type = "l", lwd = 2, col = col_sim, ylim = yl,
         xlab = "Time", ylab = "Eigenfunction value",
         main = sprintf("Factor %d, mode %d (%.0f%% of variance)", q, l, pv[l]))
    lines(time_g, Phi_hat_al[[q]][, l], lwd = 2, col = col_est)
    abline(h = 0, lty = 3, col = "grey60")
    if (q == 1 && l == 1) legend("topright", c("simulated", "estimated"),
                                 col = c(col_sim, col_est), lwd = 2, bty = "n", cex = 0.85)
  }
}
par(op)
```

### Scores: where each subject sits along each mode

The score places each subject on a given factor and mode: it is the weight of
that subject's curve on the corresponding eigenfunction, and it is where patient
heterogeneity is summarised. Each panel plots the estimated score against the
true score across subjects, for one factor and mode. A tight straight line means
the ordering and spread of subjects are recovered. The axes need not share a
scale, since loadings and scores are identified only up to a common factor, so we
report the correlation and a best-fit line.

```{r fig-scores, fig.width = 6.5, fig.height = 6.5}
op <- par(mfrow = c(Q_true, L_true), mar = c(4, 4.2, 3, 1), cex.main = 0.9)
for (q in 1:Q_true) for (l in 1:L_true) {
  xs <- sim$Zeta[[q]][, l]; ys <- Zeta_hat_al[[q]][, l]
  ok <- sd(xs) > 0 && sd(ys) > 0
  plot(xs, ys, pch = 19, col = "grey35", cex = 0.7,
       xlab = "Simulated score", ylab = "Estimated score",
       main = sprintf("Factor %d, mode %d (r = %s)", q, l,
                      ifelse(ok, sprintf("%.2f", cor(xs, ys)), "NA")))
  if (ok) abline(lm(ys ~ xs), col = col_est, lwd = 1.5)
}
par(op)
```

### Mean functions

Separately from the factors, each biomarker has its own population mean
trajectory. Each panel overlays the true mean (black) and the estimated mean
(blue) for one biomarker. These should match closely, since the mean is the
easiest part of the model to estimate.

```{r fig-means, fig.width = 9, fig.height = 4}
op <- par(mfrow = c(2, 5), mar = c(3.6, 3.6, 2.6, 1), mgp = c(2.1, 0.7, 0), cex.main = 1)
for (j in 1:p) {
  yl <- range(c(sim$mu_g[[j]], fit$list_mu_hat[[j]]))
  plot(time_g, sim$mu_g[[j]], type = "l", lwd = 2, col = col_sim, ylim = yl,
       xlab = "Time", ylab = "Mean value", main = sprintf("Biomarker %d", j))
  lines(time_g, fit$list_mu_hat[[j]], lwd = 2, col = col_est)
}
par(op)
```

## 5. Moving to real data

Replace Section 1 with your own `time_obs` and `Y` in the nested-list format.
The model works on normalised time, so first rescale every observation time to
`[0, 1]`, for example `t_scaled = (t - t_min) / (t_max - t_min)` with a common
`t_min` and `t_max` across all subjects. The returned grid `fit$time_g` and all
smooth estimates then live on `[0, 1]`, and you map them back to the original
scale with `t = t_min + t_scaled * (t_max - t_min)`. Keep the over-specified fit
of Section 2, and use Section 3 to decide how many factors and modes to retain.
Then interpret `fit$B_hat`, `fit$ppi` and `fit$factor_ppi` directly and skip the
truth alignment of Section 4.

The method scales to tens of thousands of biomarkers. For large `p`, raise
`n_cpus` to use the built-in parallelism. For a small toy like this one,
parallelism will not necessarily help and can even slow things down, because the
per-worker communication overhead outweighs the work each worker does.

If a factor comes out empty, the cause is usually weak evidence relative to the
sparsity prior. In order of impact: give each factor enough loading biomarkers
(in the simulation, raise `a_om` relative to `b_om`), increase `N`, use
`set_hyper(d_0 = 1)`, and set `bool_scale` correctly for your data.

## Reference

Jaoua, S., Temko, D. and Ruffieux, H. (2026). *A scalable Bayesian functional
factor model for high-dimensional longitudinal molecular data.*
[arXiv:2603.20844](https://arxiv.org/abs/2603.20844).
